---
title: "Using 'dollar per offer' to identify opportunities for increased investment"
author: "Joseph Ciesielski"
date: "8/16/2019"
output: 
    ioslides_presentation:
        smaller: true
        css: style.css
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.asp = 0.618, fig.align = "center")

library(tidyverse)
library(readxl)
library(here)
library(janitor)
library(broom)
library(scales)
library(randomForest)

theme_set(theme_light())

# gives date as year 1902; assume this is a data quality issue
# only 6 unique dates so I will treat this as six months worht of data

sales <- read_excel(here::here("Data Extract for Analysis - External Candidates v2.xlsx")) %>% 
  janitor::clean_names() %>% 
  mutate(sale_month = as.integer(str_sub(sale_month, 6, 7)))

sales_tidy <- sales %>% 
  group_by(sale_month, product, division, sales_channel) %>% 
  summarise_all(sum) %>% 
  ungroup() %>% 
  complete(
    sale_month, product, division, sales_channel,


    fill = list(number_of_units_offered = 0, number_of_units_sold = 0,
                total_revenue = 0)
  ) %>% 
  mutate(
    success_rate = number_of_units_sold / number_of_units_offered,
    average_sale = total_revenue / number_of_units_sold,
    dollar_per_offer = total_revenue / number_of_units_offered
  )

sales_comp <- sales_tidy %>% 
  filter(complete.cases(sales_tidy))
```

# Overview

## What are we examining?  

* `r nrow(sales_tidy %>% distinct(product))` products
* `r nrow(sales_tidy %>% distinct(sales_channel))` sales channels
* `r nrow(sales_tidy %>% distinct(division))` divisions

* Data provided on:
    * number of units offered
    * number of units sold
    * total revenue earned

## Changes over time

```{r salemonth}

sales_tidy %>%
    group_by(sale_month) %>%
    summarise(total_revenue = sum(total_revenue)) %>%
    ggplot(aes(sale_month, total_revenue)) +
        geom_line() +
        expand_limits(y = 0) +
        scale_y_continuous(labels = scales::dollar_format()) +
        labs(
            x = "month",
            y = "total revenue",
            title = "Total revenue by month"
        )
```

# How should we evaluate the success of sales? 

## Success rate 

* Sale rate is consistent across division, product, channel, month
* Accross all variables, .3% of products offered are sold

```{r successrate}

sales_tidy %>%
    filter(division != "OTHER",
           dollar_per_offer != Inf) %>%
    ggplot(aes(factor(product), success_rate)) +
        geom_point() + 
        facet_grid(division ~ sales_channel) +
        expand_limits(y = c(0, 0.004)) +
        scale_y_continuous(labels = scales::percent_format()) +
        labs(
            x = "Product",
            y = "Success rate",
            title = "Rate of successful sales"
        )
```

## Average sale

* Revenue per sale driven mostly by product
* Consistent across month and division

```{r averagesale}

sales_tidy %>%
    filter(division != "OTHER",
           dollar_per_offer != Inf) %>%
    ggplot(aes(factor(product), average_sale, color = factor(sale_month))) +
        geom_jitter() + 
        facet_grid(division ~ sales_channel) +
        scale_y_continuous(labels = dollar_format()) +
        labs(
            x = "Product",
            y = "Average sale",
            title = "Average revenue generated per sale",
            color = "Month"
        )
```

## Dollar earned per offer

* Closest possible to a 'return on investment' with the data available
* Will use this for remainder of presentation

```{r dpo}

sales_tidy %>%
    filter(division != "OTHER",
           dollar_per_offer != Inf) %>%
    ggplot(aes(factor(product), dollar_per_offer, color = factor(sale_month))) + 
        geom_jitter() +
        facet_grid(division ~ sales_channel) +
        scale_y_continuous(labels = dollar_format()) +
        labs(
            x = "Product",
            y = "Dollar earned per offer",
            title = "Dollar earned per offer",
            color = "Month"
        )

```

## More about dollar per offer

* In practice, will resemble average sale for this analysis because the sale success rate is so steady across all variables
* This metric will be more flexible if / when success rate changes
  * Incoroprates cost which average sale cannot
* Ideally, future analyses examine individual level information to understand lifetime value of customers in realtionship to the investment required to make the sale
* However, this assumes a consistent cost to offer across products, divisions, and channels
    * If costs of offer vary substantially, this analysis can be adapted by adding / estimating those costs 
* May be useful for identifying opportunities for increased investment or finding most efficient products / channels / divisions

# What influences dollar per offer? 

## Predicting dollar per offer

* I built two statistical models with four sets of variables each
* Product and sales channel provide most helpful information in predicting dollar per offer

```{r dpomod}

set.seed(2019)

sales_mod <- sales_tidy %>%
    mutate_at(vars(product, sale_month, division, sales_channel), as.factor) %>%
    filter(dollar_per_offer != Inf,
           division != "OTHER")

split <- sample(c(TRUE, FALSE), nrow(sales_mod), replace = TRUE,
                prob = c(0.7, 0.3))
train <- sales_mod[split, ]
test <- sales_mod[-split, ]


outcome <- "dollar_per_offer"
vars1 <- "product"
vars2 <- c(vars1, "sales_channel")
vars3 <- c(vars2, "division")
vars4 <- c(vars3, "sale_month")

f1 <- as.formula(paste(outcome, paste(vars1), sep = "~"))
f2 <- as.formula(paste(outcome, paste(vars2, collapse = "+"), sep = "~"))
f3 <- as.formula(paste(outcome, paste(vars3, collapse = "+"), sep = "~"))
f4 <- as.formula(paste(outcome, paste(vars4, collapse = "+"), sep = "~"))

lmod1 <- lm(f1, data = train)
lmod2 <- lm(f2, data = train)
lmod3 <- lm(f3, data = train)
lmod4 <- lm(f4, data = train)

rfmod1 <- randomForest(f1, data = train)
rfmod2 <- randomForest(f2, data = train)
rfmod3 <- randomForest(f3, data = train)
rfmod4 <- randomForest(f4, data = train)

rmse <- function(x) {

   sqrt(mean(x ^ 2))
}

sales_pred <- test %>%
    mutate(
        lpred1 = predict(lmod1, .),
        lpred2 = predict(lmod2, .),
        lpred3 = predict(lmod3, .),
        lpred4 = predict(lmod4, .),
        lresid1 = lpred1 - dollar_per_offer,
        lresid2 = lpred2 - dollar_per_offer,
        lresid3 = lpred3 - dollar_per_offer,
        lresid4 = lpred4 - dollar_per_offer,
        rfpred1 = predict(rfmod1, .),
        rfpred2 = predict(rfmod2, .),
        rfpred3 = predict(rfmod3, .),
        rfpred4 = predict(rfmod4, .),
        rfresid1 = rfpred1 - dollar_per_offer,
        rfresid2 = rfpred2 - dollar_per_offer,
        rfresid3 = rfpred3 - dollar_per_offer,
        rfresid4 = rfpred4 - dollar_per_offer,
    )

sales_pred %>%
    summarise_at(vars(contains("resid")), rmse) %>%
    gather(model, resid) %>% 
    separate(model, c("type", "vars"), sep = "resid") %>%
    ggplot(aes(vars, resid, color = type, group = type)) +
        geom_line() +
        expand_limits(y = 0) +
        scale_color_discrete(labels = c("linear model", "random forest")) +
        scale_x_discrete(labels = c("product", "product + channel",
                                    "product + channel +\ndivision",
                                    "product + channel +\ndivision + month")) +
        labs(
            title = "Product and sales channel provide most useful information",
            x = "Predictors",
            y = "RMSE", 
            color = "Model",
            subtitle = "Root mean squared error of each model when applied to unseen test data"
        )

```

# What are the most profitable products? 

## Most revenue driven by a few products

* Products 3 and 4 are the highest selling, offered most
* Channels Z and X generate the most revenue

```{r div1}

sales_tidy %>%
    filter(division != "OTHER") %>%
    mutate(product = fct_lump(factor(product), n = 7, w = total_revenue)) %>%
    ggplot(aes(factor(product), total_revenue, fill = sales_channel)) +
        geom_col(position = "dodge") +
        facet_wrap(~ division, ncol = 1) +
        scale_y_continuous(labels = dollar_format()) +
        labs(
            x = "Product",
            y = "Total revenue",
            fill = "Sales channel",
            title = "Total revenue by channel and division for top products"
        )

```

## Revenue vs efficiency

* Channel Y tends to offer a higher return on investment
* Products 11 and 12 also have a higher dollar per offer but generate fairly low amounts of revenue

```{r}

sales_tidy %>%
    mutate(product = fct_lump(factor(product), n = 7, w = total_revenue)) %>%
    filter(division != "OTHER") %>%
    ggplot(aes(factor(product), average_sale, color = sales_channel)) +
        geom_boxplot() +
        facet_wrap(~ division, ncol = 1) +
        labs(
            title = "Top selling product not necessarily the most profitable",
            subtitle = "Range of dollar per offer for products by channel and division",
            x = "Product",
            y = "Dollar per offer",
            color = "Sales channel"
        )


```

# Next steps 

## Recommendations 

* Explore opportunities for increasing sales in Channel Y and for Products 11 and 12
  * These tend to offer the highest return on investment even though they represent a relatively small proportion of revenue
* Investiagte whether Product 3 and Channel Y should continue to represent the highest proportion of revenue
* Best to explore these questions are randomized experiments
  * Establish key metrics (dollar per offer or something else)
  * Increase / decrease offers in one or more of the above areas and measure if revenue increases or decerases

## Outstanding questions

* What is the cost of each offer? Is it consistent or variable? 
    * If variable, need to incorporate this information before making recommendations
* Why is success rate so consistent? 
* Conduct deeper analysis using individual level information about customers
    * What does it cost to acquire a customer?
    * Can we predict a customer's lifetime value?
    * Do some products / channels / divisions have better customer retention? 

